Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-30 manuscript/manuscript.Rmd
added stable_points.py

bibliography: bibliography.bib
#biblio-style: spphys

@@

As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
There are a variety of fixed points for the equations---around 7 or 8 for wide ranges of parameter values.
Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.

Throughout most of this paper we will fix the parameters of the model according to Table \ref{tab:tbd}.
At these values, the fixed points of the model are:
\begin{align*}
  TBD
\end{align*}
The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(TBD)$ is stable against perturbations in all directions.
We will thus use it as a starting point for our simulations.

@@

Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
Specifically, we parametrize $\pi$ using a neural network with two hidden layers of 64 neurons.

We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.
Here we have focused on a single one of these for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
In practice, further improvements can be expected by a careful selection of the optimization algorithm.

@@

Within control theory, the usual setup is one where we use as much information from the model as possible in order to derive an optimal solution.
**(Wax poetic a bit about Bellmann eq. approaches here?)**

The classical sustainable fishery management approaches summarized in Sec. \ref{sec:classical} are model-based controls.
As we saw in that section, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.

More generally, there are many situations in which the exact model of the system is not known or not tractable.
This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
In this case, it is clear, model-based controls run a grave danger of mismanaging the system.

Reinforcement learning, on the other hand, can provide a model-free approach to control theory.
While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
This provides more flexibility to use RL in instances where the model of the system is not accurately known.
In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
(Note that there are approaches to perform model-based reinforcement learning. 
While we will not focus on these in this paper they are surveyed in, e.g., [@moerland2023model; @polydoros2017survey].)

**Mention CL here? Where solutions to other related problems in the curriculum are used to build solutions to the target problem.**

@@

\textbf{Parameter values used.}
The model's dynamic parameters were chosen as follows
\begin{align*}
  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
\end{align*}
Moreover, the variances for the stochastic terms were chosen as
$$
  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
$$
The threshold values used to determine an early episode end were chosen as
$$
  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
$$
where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).

@@

# References

>>>>

bibliography: bibliography.bib
biblio-style: spphys

@@

As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
Fig. \ref{fig:tbd} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.

Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
At these values, the fixed points of the model are:\footnote{
We have included only three decimal digits for readability.
These fixed points may be generated by running the ‘‘stable\_points.py’’ script in [@repo].
}
\begin{align*}
  (X,Y,Z) \approx &(0.839, 0.054, 0.377), \quad 
    (0.048, 0.774, 0.480), \quad
    (0.359, 0.491, 0.509), \\
    &(0.666, 0.666, 0.000),\quad
    (0.899, 0.000, 0.303), \quad
    (0.000, 0.818, 0.457), \\
    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
\end{align*}
The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
We will thus use it as a starting point for our simulations.

@@

Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
We use the *proximal policy optimization (PPO)* algorithm to optimize $\pi$.
Within the RL literature there is a wealth of algorithms from which to choose from, each with its pros and cons.
Here we have focused on a single one of these for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
In practice, further improvements can be expected by a careful selection of the optimization algorithm.

@@

Within control theory, the classical setup is one where we use as much information from the model as possible in order to derive an optimal solution.
Here, one may find a vast literature on model-based methods to attain optimal, or near-optimal, control (see, e.g., [@zhang2019near; @sethi2019optimal; @anderson2007optimal]).

The classical sustainable fishery management approaches summarized in Text Box 1, for instance, are model-based controls.
As we saw there, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.

There are many situations, however, in which the exact model of the system is not known or not tractable.
This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
In this case, it is clear, model-based controls run a grave danger of mismanaging the system.

Reinforcement learning, on the other hand, can provide a model-free approach to control theory.
While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
This provides more flexibility to use RL in instances where the model of the system is not accurately known.
In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
(For recent surveys of model-based reinforcement learning, which we do not focus on here, see [@moerland2023model; @polydoros2017survey].)

@@

\textbf{Parameter values used.}
The model's dynamic parameters were chosen as follows
\begin{align}
\label{eq:param vals}
\begin{split}
  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
\end{split}
\end{align}
Moreover, the variances for the stochastic terms were chosen as
$$
  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
$$
The threshold values used to determine an early episode end were chosen as
$$
  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
$$
where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).

@@
